---
title: "Pathway workshop - Data preparation"
author: "Mark Ziemann & Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
theme: cosmo
---

Source: https://github.com/markziemann/pathway-workshop

## Introduction

```{r,libs}

#knitr::opts_chunk$set(dev = 'svg') # set output device to svg

suppressPackageStartupMessages({
  library("getDEE2")
  library("DESeq2")
  library("kableExtra")
  library("eulerr")
  library("gplots")
  library("mitch")
  library("fgsea")
  library("RhpcBLASctl")
  library("parallel")
  library("beeswarm")
  library("vioplot")
})

RhpcBLASctl::blas_set_num_threads(1)

```

For this guide I will be using bulk RNA-seq data from a previous study, which is
deposited at NCBI GEO and SRA under accession numbers: GSE55123 and SRP038101 (Lund et al, 2014).
The experiment is designed to investigate the effect of Azacitidine treatment on AML3 cells.

The raw data have been processed by the DEE2 project, and the summary gene expression counts are
available at the dee2.io website, and programmatically with the getDEE2 bioconductor package
(Ziemann et al, 2019).

The gene counts have also been deposited to the `/data` folder in the `example.Rdata` file in case
the DEE2 resource becomes unavailable.
To import it, use the command: `load("data/example.Rdata")`

Alternatively, you could fetch data from another resource like NCBI GEO, Zenodo or from the host
storage drive.

```{r,fetch}

mdat <- getDEE2Metadata("hsapiens")

# get sample sheet
ss <- subset(mdat,SRP_accession=="SRP038101")

# fetch the whole set of RNA-seq data
x <- getDEE2("hsapiens",ss$SRR_accession , metadata=mdat, legacy=TRUE)
mx <- x$GeneCounts
rownames(mx) <- paste(rownames(mx),x$GeneInfo$GeneSymbol)
dim(mx)

# aza no filtering
ss$trt <- grepl("Treated",ss$Experiment_title)

ss %>%
  kbl(caption="sample sheet for Aza treatment in AML3 cells") %>%
  kable_paper("hover", full_width = F)

```

## Data quality control

QC is important, even if you are using public transcriptome data.
For RNA-seq it is a good idea to show the number of reads for each sample.

```{r,counts1}

colSums(mx)

par(mar=c(5,7,5,1))
barplot(rev(colSums(mx)),horiz=TRUE,las=1,main="number of reads per sample in SRP038101")

```

Now make a MDS plot.

```{r,mds1}

mds <- cmdscale(dist(t(mx)))

# expand plot area
XMIN=min(mds[,1])*1.3
XMAX=max(mds[,1])*1.3
YMIN=min(mds[,2])*1.3
YMAX=max(mds[,2])*1.3

cols <- as.character(grepl("Treated",ss$Experiment_title))
cols <- gsub("FALSE","lightblue",cols)
cols <- gsub("TRUE","pink",cols)
plot(mds, xlab="Coordinate 1", ylab="Coordinate 2",
  xlim=c(XMIN,XMAX),ylim=c(YMIN,YMAX),
  pch=19,cex=2,col=cols, main="MDS plot")
text(cmdscale(dist(t(mx))), labels=colnames(mx) )

```

## Differential expression analysis

We will be using DESeq2 for DE analysis.

The count matrix is prefiltered using a detection threshold of 10 reads per sample across all samples.
All genes that meet the detection threshold will comprise the background list.

We will also repeat this without a detection threshold.

The first 6 rows of the count matrix are shown.

```{r,de1}

mxf <- mx[which(rowMeans(mx)>=10),]
dim(mxf)

head(mxf,6) %>%
  kbl(caption="Count matrix format") %>%
  kable_paper("hover", full_width = F)

dds <- DESeqDataSetFromMatrix(countData = mxf , colData = ss, design = ~ trt )
res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <-cbind(as.data.frame(z),mxf)
def <-as.data.frame(zz[order(zz$pvalue),])

head(def,10) %>%
  kbl(caption="Top DE genes for Aza treatment") %>%
  kable_paper("hover", full_width = F)

```

Make a smear plot.

```{r,smearplot}

sigf <- subset(def,padj<=0.01)

DET=nrow(mxf)
NSIG=nrow(sigf)
NUP=nrow(subset(sigf,log2FoldChange>0))
NDN=nrow(subset(sigf,log2FoldChange<0))

HEADER=paste(DET,"detected genes;",NSIG,"w/FDR<0.01;",NUP,"up;",NDN,"down")

plot(log10(def$baseMean) ,def$log2FoldChange,
  cex=0.6,pch=19,col="darkgray",
  main="5-azacitidine treatment in AML3 cells",
  xlab="log10(basemean)",ylab="log2(fold change)")

points(log10(sigf$baseMean) ,sigf$log2FoldChange,
  cex=0.6,pch=19,col="red")

mtext(HEADER)

```

In the next sections I will run enrichment analysis with over-representation test
and compare it to functional class scoring.

## Get update gene symbols

Let's see if many of the gene symbols have changed.

Comparing Ensembl release 90 to GENCODE Release 49 (GRCh38.p14) which corresponds to Ensembl 115 (Sept 2025).

So v90 had 58k genes and v115 has 86k which is a big increase.
The v90 has 1609 genes that have been removed an v115 has 29676 that have been added.
There are 56693 common.

From these common ones, 20456 have different gene names.
I think the approach to naming these putative genes has changed a lot.
Lots of CXorfXX genes have changed, not just AL358472.5 type names.

In the next chunk, I'll show a few lines of the older source gene names.

There is also an Euler diagram of the gene symbol repertoires.

```{r,genesymbolupdate,fig.height=3,fig.width=5}

gt <- read.table("https://dee2.io/data/hsapiens/hsa_gene_info.tsv",header=TRUE)
head(gt)

gtnew <- read.table("ref/gencode.v49.genenames.tsv")

v1 <- list("Ens90"=gt$GeneID, "Ens115"=gtnew$V1)
plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))

```

```{r,genesymbolupdate2,fig.height=3,fig.width=5}

message("Number of unique genes in the original gene table:")
length(gt$GeneID)

message("Number of unique genes in the new gene annotation set:")
length(gtnew$V1)

message("Number of shared gene IDs between new and old annotations:")
length(intersect(gt$GeneID,gtnew$V1))

message("Number of gene IDs specific to the old annotation:")
length(setdiff(gt$GeneID,gtnew$V1))

message("Number of gene IDs specific to the new annotation:")
length(setdiff(gtnew$V1,gt$GeneID))

message("Here is the mapping table for old-new gene symbols:")
gtm <- merge(gt,gtnew,by.x="GeneID",by.y="V1")

message("Number of rows:")
nrow(gtm)

message("How many of these gene symbols are the same?")
table(gtm$GeneSymbol == gtm$V2)

```

## Save data

```{r,save}

dir.create("data")

detsv <- cbind(rownames(def),def)
colnames(detsv)[1] <- "GeneID"
write.table(detsv,"data/de.tsv",sep="\t",row.names=FALSE)
save.image("data/de.Rdata")
saveRDS(def,"data/de.Rds")

```

## Session information

```{r,session}

sessionInfo()

```
