---
title: "Pathway analysis powered with R/bioconductor"
author: "Mark Ziemann & Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
theme: cosmo
---

Source: https://github.com/markziemann/pathway-workshop

## Introduction

```{r,libs}

#knitr::opts_chunk$set(dev = 'svg') # set output device to svg

suppressPackageStartupMessages({
  library("getDEE2")
  library("DESeq2")
  library("kableExtra")
  library("eulerr")
  library("gplots")
  library("mitch")
  library("fgsea")
  library("RhpcBLASctl")
  library("parallel")
  library("beeswarm")
  library("vioplot")
})

RhpcBLASctl::blas_set_num_threads(1)

```

Here we demonstrate how to do over-representation analysis (ORA) and functional class scoring (FCS).

## Load DE data

For information on the dataset we are using, see the script called "dataprep.Rmd".

```{r,load}

de <- readRDS("data/de.Rds")

head(de) %>%
  kbl(caption="To DEGs for Aza treatment in AML3 cells") %>%
  kable_paper("hover", full_width = F)

```

Gene sets were obtained from Reactome.

```{r,reactome}

gs <- gmtPathways("ref/ReactomePathways_2025-09-02.gmt")

head(gs,3)

summary(unlist(lapply(gs,length)))

```

## ORA: Filter genes

Here we select up and down-regulated genes as well as the background.
The background was previously selected based on the presence of an
average of 10 reads per sample.

```{r,filter}

up <- rownames(subset(de, padj < 0.01 & log2FoldChange > 0))
up <- unique(sapply(strsplit(up," "),"[[",2))
length(up)

dn <- rownames(subset(de, padj < 0.01 & log2FoldChange < 0))
dn <- unique(sapply(strsplit(dn," "),"[[",2))
length(dn)

# background selection was done earlier
#mxf <- mx[which(rowMeans(mx)>=10),]
dim(de)
bg <- rownames(de)
bg <- unique(sapply(strsplit(bg," "),"[[",2))
length(bg)

```

Now we write the gene lists to text files that we can use for ORA with web tools like iDEP.

```{r,write_gene_lists}

writeLines(up,"data/up.txt")
writeLines(dn,"data/down.txt")
writeLines(bg,"data/background.txt")

```

## ORA: run the test with reactome gene sets

Separate tests for up and down.

First, up.

```{r,ora_up1}

ora_up <- fora(pathways=gs, genes=up, universe=bg, minSize = 5, maxSize = Inf)
nrow(ora_up)
message("number of up-regulated gene sets with FDR<0.01")
nrow(subset(ora_up,padj<0.01))

head(ora_up,10) %>%
  kbl(caption="Top upregulated pathways with ORA (significance)") %>%
  kable_paper("hover", full_width = F)

ora_up <- ora_up[order(-ora_up$foldEnrichment),]

head(ora_up,10) %>%
  kbl(caption="Top upregulated pathways with ORA (fold enrichment)") %>%
  kable_paper("hover", full_width = F)

```

Next, down.

```{r,ora_dn1}

ora_dn <- fora(pathways=gs, genes=dn, universe=bg, minSize = 5, maxSize = Inf)
nrow(ora_dn)
message("number of down-regulated gene sets with FDR<0.01")
nrow(subset(ora_dn,padj<0.01))

head(ora_dn,10) %>%
  kbl(caption="Top downregulated pathways with ORA (significance)") %>%
  kable_paper("hover", full_width = F)

ora_dn <- ora_dn[order(-ora_dn$foldEnrichment),]

head(ora_dn,10) %>%
  kbl(caption="Top dnregulated pathways with ORA (fold enrichment)") %>%
  kable_paper("hover", full_width = F)

```

Barplot

```{r,ora_barplot}

npw=15
ora_up_vals <- head(ora_up$foldEnrichment,15)
names(ora_up_vals) <- head(ora_up$pathway,15)

ora_dn_vals <- head(ora_dn$foldEnrichment,15)
names(ora_dn_vals) <- head(ora_dn$pathway,15)

myvals <- c(ora_dn_vals,rev(ora_up_vals))

names(myvals) <- ifelse(nchar(names(myvals)) > 53, paste0(strtrim(names(myvals), 50), '...'), names(myvals))

par(mar=c(5.1,25.1,4.1,2.1))

mycols <- c( rep("blue",length(ora_dn_vals)) , rep("red",length(ora_up_vals)) )

barplot(myvals, col=mycols,
  horiz=TRUE,las=2,xlab="Fold enrichment",main="Effect of Aza treatment")
mtext("Red=up, Blue=down")

par(mar=c(5.1,4.1,4.1,2.1))


```


## Session information

```{r,session}

sessionInfo()

```
