---
title: "Pathway analysis powered with R/bioconductor"
author: "Mark Ziemann & Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

Source: https://github.com/markziemann/pathway-workshop

## Introduction

```{r,libs}

#knitr::opts_chunk$set(dev = 'svg') # set output device to svg

suppressPackageStartupMessages({
  library("getDEE2")
  library("DESeq2")
  library("kableExtra")
  library("eulerr")
  library("gplots")
  library("mitch")
  library("fgsea")
  library("RhpcBLASctl")
  library("parallel")
  library("beeswarm")
  library("vioplot")
})

RhpcBLASctl::blas_set_num_threads(1)

```

Here we demonstrate how to do over-representation analysis (ORA) and functional class scoring (FCS).

## Load DE data

For information on the dataset we are using, see the script called "dataprep.Rmd".

```{r,load}

de <- readRDS("data/de.Rds")

head(de) %>%
  kbl(caption="To DEGs for Aza treatment in AML3 cells") %>%
  kable_paper("hover", full_width = F)

```

Gene sets were obtained from Reactome.

```{r,reactome}

gs <- gmtPathways("ref/ReactomePathways_2025-09-02.gmt")

head(gs,3)

summary(unlist(lapply(gs,length)))

```

## ORA: Filter genes

Here we select up and down-regulated genes as well as the background.
The background was previously selected based on the presence of an
average of 10 reads per sample.

```{r,filter}

up <- rownames(subset(de, padj < 0.01 & log2FoldChange > 0))
up <- unique(sapply(strsplit(up," "),"[[",2))
length(up)

dn <- rownames(subset(de, padj < 0.01 & log2FoldChange < 0))
dn <- unique(sapply(strsplit(dn," "),"[[",2))
length(dn)

# background selection was done earlier
#mxf <- mx[which(rowMeans(mx)>=10),]
dim(de)
bg <- rownames(de)
bg <- unique(sapply(strsplit(bg," "),"[[",2))
length(bg)

```

Now we write the gene lists to text files that we can use for ORA with web tools like iDEP.

```{r,write_gene_lists}

writeLines(up,"data/up.txt")
writeLines(dn,"data/down.txt")
writeLines(bg,"data/background.txt")

```

## ORA: run the test with reactome gene sets

The goal is to understand whether certain gene pathways are enriched at a higher level in the
DE list as compared to the background list of detected genes.

We are running separate tests for up and down DEGs.

First, up.

```{r,ora_up1}

ora_up <- fora(pathways=gs, genes=up, universe=bg, minSize = 5, maxSize = Inf)
nrow(ora_up)
message("number of up-regulated gene sets with FDR<0.01")
nrow(subset(ora_up,padj<0.01))

head(ora_up,10) %>%
  kbl(caption="Top upregulated pathways with ORA (significance)") %>%
  kable_paper("hover", full_width = F)

ora_up <- ora_up[order(-ora_up$foldEnrichment),]

head(ora_up,10) %>%
  kbl(caption="Top upregulated pathways with ORA (fold enrichment)") %>%
  kable_paper("hover", full_width = F)

```

Next, down.

```{r,ora_dn1}

ora_dn <- fora(pathways=gs, genes=dn, universe=bg, minSize = 5, maxSize = Inf)
nrow(ora_dn)
message("number of down-regulated gene sets with FDR<0.01")
nrow(subset(ora_dn,padj<0.01))

head(ora_dn,10) %>%
  kbl(caption="Top downregulated pathways with ORA (significance)") %>%
  kable_paper("hover", full_width = F)

ora_dn <- ora_dn[order(-ora_dn$foldEnrichment),]

head(ora_dn,10) %>%
  kbl(caption="Top dnregulated pathways with ORA (fold enrichment)") %>%
  kable_paper("hover", full_width = F)

```

Barplot

```{r,ora_barplot}

npw=15
ora_up_vals <- head(ora_up$foldEnrichment,15)
names(ora_up_vals) <- head(ora_up$pathway,15)

ora_dn_vals <- head(ora_dn$foldEnrichment,15)
names(ora_dn_vals) <- head(ora_dn$pathway,15)

myvals <- c(ora_dn_vals,rev(ora_up_vals))

names(myvals) <- ifelse(nchar(names(myvals)) > 53, paste0(strtrim(names(myvals), 50), '...'), names(myvals))

par(mar=c(5.1,25.1,4.1,2.1))

mycols <- c( rep("blue",length(ora_dn_vals)) , rep("red",length(ora_up_vals)) )

barplot(myvals, col=mycols,
  horiz=TRUE,las=2,xlab="Fold enrichment",main="Effect of Aza treatment")
mtext("Red=up, Blue=down")

par(mar=c(5.1,4.1,4.1,2.1))

```

## Functional class scoring

We will rank the genes based on the "stat" value from DESeq2, which is the Wald test statistic.

First step is to remove redundant gene symbols (there are 4).

```{r,fcs1,message = FALSE}

de2 <- de

de2$gene <- sapply(strsplit(rownames(de)," "),"[[",2)

which(duplicated(de2$gene))

de2ag <- aggregate(de2, stat ~ gene, FUN=sum)

destat <- de2ag$stat
names(destat) <- de2ag$gene

head(destat)

fcs_res <- fgsea(pathways = gs,
                  stats    = destat,
                  minSize  = 5,
                  maxSize  = 50000,
                  nPermSimple = 10000,
                  nproc = 1)

fcs_res <- fcs_res[order(fcs_res$padj),]

```

Now let's take a look at the top results based on statistical significance.

```{r,fcs2}

head( subset(fcs_res,ES>0) , 10) %>%
  kbl(caption="Top upregulated pathways with FCS (significance)") %>%
  kable_paper("hover", full_width = F)

head( subset(fcs_res,ES<0) , 10) %>%
  kbl(caption="Top downregulated pathways with FCS (significance)") %>%
  kable_paper("hover", full_width = F)

```

Now let's take a look at the results based on ES after removal of nonsignificant sets.

```{r,fcs3}

sig <- subset(fcs_res,padj<0.01)

fcs_up <- tail(sig[order(sig$ES),],15)

fcs_up %>%
  kbl(caption="Top upregulated pathways with FCS (ES prioritisation)") %>%
  kable_paper("hover", full_width = F)

fcs_dn <- head(sig[order(sig$ES),],15)

fcs_dn %>%
  kbl(caption="Top downregulated pathways with FCS (ES prioritisation)") %>%
  kable_paper("hover", full_width = F)

```

Now make a barplot.

```{r,fcs_barplot}

fcs_vals <- c(fcs_dn$ES, fcs_up$ES)
names(fcs_vals) <- c(fcs_dn$pathway, fcs_up$pathway)

names(fcs_vals) <- ifelse(nchar(names(fcs_vals)) > 53, paste0(strtrim(names(fcs_vals), 50), '...'), names(fcs_vals))

par(mar=c(5.1,24.1,4.1,2.1))

barplot(fcs_vals, horiz=TRUE,las=2,xlab="ES",main="Effect of Aza treatment")

par(mar=c(5.1,4.1,4.1,2.1))

```

## Compare ORA and FCS approaches

```{r,euler}

ora_sig_up <- subset(ora_up,padj<0.01)$pathway
ora_sig_dn <- subset(ora_dn,padj<0.01)$pathway

fcs_sig_up <- subset(fcs_res,ES>0 & padj<0.01)$pathway
fcs_sig_dn <- subset(fcs_res,ES<0 & padj<0.01)$pathway

v1 <- list("ORA up"=ora_sig_up, "ORA down"=ora_sig_dn,
  "FCS up"=fcs_sig_up, "FCS down"=fcs_sig_dn)

plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))

```

Set analysis.

```{r,set_analysis1}

message("Common upregulated")
intersect(ora_sig_up, fcs_sig_up)

message("Common downregulated")
intersect(ora_sig_dn, fcs_sig_dn)

```

Enrichment plot.

```{r,eplot}

mypw <- gs[which(names(gs) %in% fcs_dn$pathway)]

plotGseaTable(mypw, destat, fcs_res, gseaParam = 0.5)

```

## Session information

```{r,session}

sessionInfo()
save.image("data/session2.Rdata")

```
